master thesis

### 2018-03-25 Paper - Latex/Paper - Skizze - v4.tex
Proofreading

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %--------------------------------------------------------------------------------------------------- 
  \section{Motivation}
    The aim of this thesis is the application of maximum parsiomony algorithms to non-binary 
      trees and very large datasets. In particular, the example find the 'Origins of Parasitism' 
      throughout the Eukaryota Tree of Life. \\
    For these large phylogenetic synthesis trees, however, ancestral state reconstruction has so far 
      only been done for Bacteria and Archaea for binary traits by Goberna and Verdú \cite{Goberna2015}.
      However, this differs from eukaryotes in the sense that complex traits such as parasitism depend
      on more than one gene. \\
    \todo{Warum ist das relevant? Was könnte man dann tun?}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %--------------------------------------------------------------------------------------------------- 
  \section{Background}
    For about 50 years, people have been working on ancestral state reconstruction. The first paper 
      is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
      \cite{Camin1965}. Different methods have been developed and the question is which method is the 
      most suitable for the problem at hand: The ancestral state reconstruction for a huge 
      multifurcated tree with \colorbox{red}{binary/two discrete} states. \\
    Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
    The first is maximum parsimony: explain the current state with the least number of state changes 
      between the child and his ancestor. \\
    The other class she presents describes modeling the character evolution as a stochastic process and 
      using the likelihoods to compute the possible ancestral character states. This is generally done 
      with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

@@

    We will test the existing maximum parsimony algorithms Fitch \cite{Fitch1971} and Sankoff 
      \cite{Sankoff1975} for this task and estimate their predictive power. The present tree structure 
      of OTL is not binary. A tree is multifurcated if each node has multiple ($n >= 3$) children 
      \cite{Felsenstein2003}. \\
    Maximum parsimony methods are developed for phylogenies, which are usually depicted as binary trees.
      Parsimonious in phylogeny refers to favoring the tree that needs the least evolutionary change 
      to explain the observed data. In our case, it is about the change of states 'is free-living' or 
      'is parasitic'. \\
      % Parsimonious in this context means to minimize the number of transitions between species states 
      % (free-living or parasitic). 
    The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
      package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
      method and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
    This achieves that we can predict ancestral states and unknown states of living species for large 
      non-binary relatives trees. \\

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  As stated in the introduction, in this thesis, a maximum parsimony algorithm is applied to the 
    whole tree of life to obtain an ancestral state reconstruction of free-living versus parasite 
    states. \\
  So far, these reconstructions have been made mainly on binary trees with better data availability. 
    Therefore, a simulation is first performed to evaluate existing algorithms and decide how they 
    may be adapted to our given problem. This is to perform the ancestral state reconstruction for a 
    multifurcated (non-binary) tree using binary states. \\
  Accordingly, in addition to the necessary data sets (GloBI, OTL), the chosen algorithm and the 
    evaluation of its results, this chapter also deals with the previously performed simulation and 
    the evaluation of the various algorithms and their parameters. \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc the simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulted algorithm on the original data.
      (7) Evalute and interprete results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Description of data sets}
    Two types of data are needed for an ancestral state reconstruction: a tree and information about 
      the states. \\
    For the Tree Open Tree of Life (OTL) is used \cite{Hinchliff2015} and for the state information 
      the Global Biotic Interaction database (GloBI) is taken \cite{Poelen2014}. \\
    % These are presented below.

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 0mm, clip, width=0.45\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{60~\% Free-living - 40~\% Parasites \\ red: parasites, blue: free-living, \\ the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

@@

    Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
      use case here. Following the principle of the simpler model first. \\
    \anmerkungstext{Nochmal darstellen warum und selbst wenn sie fully sufficient wäre, heißt es ja 
      nicht, dass man sie nehmen muss. Also: motiviere! (Renard)} \\
    Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
      previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
      \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
    \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
      Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
    \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times trough the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      \begin{wrapfigure}{r}{0.5\textwidth}
        \centering
        \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
        \caption{Fitch algorithm for binary trees. \\
          The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
          root node) consists of three sets, where the last set is the final one (bold). \\
          From the second internal node (seen from the root node) there are several possibilities to 
          create the second and third set.}
        \label{fig: binary Fitch}
      \end{wrapfigure}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{real data analysis}
    \todo{Evaluation section in methods?}
    \begin{itemize}
      \item Import tree
      \item Import interactions
      \item run castor algorithm / and others?
      \item interprete results (leave-100-out cross validation)
    \end{itemize}
    We then evaluated the results in two ways. On the one hand, we have examined the results of some 
      subtrees for their credibility with the help of biological background knowledge. On the other 
      hand, we performed a leave-100-out cross validation.

@@

  We then evaluated the results in two ways. On the one hand, we have examined the results of some 
    subtrees for their credibility with the help of biological background knowledge (subsection 
    \ref{sec:results - biological view}). On the other hand, we performed a leave-100-out cross validation
    (Subsection \ref{sec:results - cross validation}) and predicted approximately 98.17~\% states 
    correct.

@@

    We also have for a first overview we collected for each node its number of children (degree $-1$), 
      and plotted this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    \begin{figure}
      \centering
      \begin{subfigure}[b]{0.59\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~}
      \end{subfigure}
      \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. Higher multifurcation than 30 has been cut off. \\ light blue: binary; \\ dark blue: multifurcation}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree. \\
        For each node, the number of children (degree $-1$) was collected. A node is multifurcated 
        if it has more than two children, so we deducted two from each number of children. We have 
        plotted this in two histograms: number of children $-2$ on the x-axis with log scale and the 
        number of nodes with this amount on the y-axis.}
      \label{fig:childrenOfNodes}
    \end{figure}
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it discribes the number of nodes 
      which we have lost from the real (binary) phylogenetic tree. \\
    It can be recognized that we are very far from a binary tree. \\

@@

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            \bfseries Model / Taxa & \bfseries Kingdom & \bfseries Phylum & \bfseries Class & \bfseries Order \\
            \hline \hline
            multifurc $\sim$ taxa & 8273333 & \cellcolor{green!15}7937828 & \cellcolor{green!20}7842157 & \cellcolor{green!30}7644249 \\
            multifurc $\sim$ taxa & 8257680 & \cellcolor{green!15}7922207 & \cellcolor{green!20}7826490 & \cellcolor{green!35}7574154 \\
            \hline
            multifurc $\sim$ taxa + depth & 8273318 & \cellcolor{green!15}7934322 & \cellcolor{green!20}7839364 & \cellcolor{green!35}7539999 \\
            multifurc $\sim$ taxa + max.height & \cellcolor{green!15}7993515 & \cellcolor{green!25}7749121 & \cellcolor{green!30}7661817 & \cellcolor{green!40}7416211 \\
            multifurc $\sim$ taxa + min.height & 8251211 & \cellcolor{green!20}7875521  & \cellcolor{green!25}7778327 & \cellcolor{green!35}7516883 \\
            multifurc $\sim$ taxa + mean.height & \cellcolor{green!20}7825417 & \cellcolor{green!30}7644249 & \cellcolor{green!35}7572474 & \cellcolor{green!45}7340741 \\
            \hline
            multifurc $\sim$ taxa * depth & 8235932 & \cellcolor{green!20}7836755 & \cellcolor{green!25}7757688 & \cellcolor{green!45}7383808 \\
            multifurc $\sim$ taxa * max.height & \cellcolor{green!15}7963438 & \cellcolor{green!30}7693555 & \cellcolor{green!30}7614820 & \cellcolor{green!45}7335338 \\
            multifurc $\sim$ taxa * min.height & 8214030 & \cellcolor{green!20}7808940 & \cellcolor{green!30}7690618 & \cellcolor{green!45}7336627\\
            multifurc $\sim$ taxa * mean.height & \cellcolor{green!25}7768360 & \cellcolor{green!35}7536296 & \cellcolor{green!50}7484953 & \cellcolor{green!50}7206369 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BIC (Bayesian information criterion) values of the multifurcation models. \\
          These models were created with the R function \textit{glm()} and compared with the 
            \textit{BIC()} function. This results in the listed BIC values.}
        \label{table:BIC multifurcation} 
      \end{table}
      % * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\
      Within the every complexity class it can be seen that the mean height gives the best additional factor. \\
      Despite higher complexity, the BIC values are getting smaller from model to model, meaning that 
        the finest model available here is also the best one of these. Lower taxa than orders (eg 
        family) were computationally too expensive to calculate. \\
      The model \textit{multifurc $\sim$ order * mean.height} turns out to be the best of our models, 
        whereby it is possible that eg \textit{multifurc $\sim$ family * mean.height} is better.

@@

      Next to the problem of the multifurcation of the tree is the less of interaction data we have for 
        the species. For the ancestral state reconstruction, we need information about the states 
        (free-living or parasite) in the leaf nodes. \\
      The eukaryotic synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we got 51,337 distinct free-living species and 47 332 
        distinct parasite species for the whole tree of life. From the Eukaryota we could determine 
        25,992 and 34,879 species as parasites and free-living. With 2,293,463 leaf nodes we still have 
        about 97.34~\% unknown leaf nodes. In the discussion, section 
        \ref{sec:discussion - data situation}, we will talk about ways to extend this data. \\

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            \bfseries Model / Taxa & \bfseries Kingdom & \bfseries Phylum & \bfseries Class & \bfseries Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & \cellcolor{green!45}484681 \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BIC (Bayesian information criterion) values unknown state information models. \\
          These models were created with the R function \textit{glm()} and compared with the 
            \textit{BIC()} function. This results in the listed BIC values.}
        \label{table:BIC unknown information} 
      \end{table}
      It also follows from this table that the most complex model is the best. In general, the BIC 
        values are smaller than those of the multifurcation models. The modeling here is thus better. \\
      Again, the calculation of finer models (eg order or family) was too expensive. \\

@@

    As presented, we compare two methods in our simulation to their prediction accuracy: Fitch and 
      Sankoff. \\
    Figure \ref{fig:influence of unknown data} shows the results of examining various parameters.
    \begin{figure}[h!]
      \centering
      \includegraphics[trim = 0mm 100mm 0mm 0mm, clip, width=\textwidth]{Figures/simulation_evaluation_1.png}
    \end{figure}
    ~ \\ ~ \\ ~ \\
    \begin{figure}
      \centering
      \includegraphics[trim = 0mm 0mm 0mm 150mm, clip, width=\textwidth]{Figures/simulation_evaluation_1.png}
      \caption{Influence of unknown data to prediction. \\
        The first column describes the distributions of free-livings and parasites with a given 
          threshold for the respective simulations to the right. \\
        The middle column investigates the influence of the unknown states, the right the influence of
          the strength of the multifurcation. \\
        The y-axes indicate the percentage of correctly predicted states (including known states). On 
          the x-axis the percentage of forgotten states or missing internal nodes. \\
        Each point corresponds to the average of one hundred simulations, each with 10,000 leaf nodes. \\
        For the middle column we set the strength of the multifurcations to 0.95~\% similar to the 
          real data and in the right column the amount of the unknowns to 0.95~\% also similar to the 
          real data.}
      \label{fig:influence of unknown data}
    \end{figure}
    First, we tested different trasition probability distributions including threshold (first column). 
      It can be observed that the more balanced the percentages of free-livings and parasites, the 
      worse the prediction of the algorithms (second and third column). \\
    It can be seen that both algortithms are always over 50~\% prediction and therefore better than 
      guessing. Moreover, they are usually close to each other, with Sankoff always makes better 
      predictions except for equally distributed states as Fitch. \\
    On the other hand, we examined the influence of missing internal nodes (ridge of multifurcation) 
      and missing leaf node information (unknown leaf nodes). \\
    Both factors have a relatively linear influence on the Sankoff method. Fitch, on the other hand, 
      breaks significantly in his prediction from about 70\% unknown leaf nodes or mutifurcation. \\

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Results of the real data analysis created with Sankoff}
    This section is about evaluating the prediction of real data using the Sankoff method. \\
    It is divided into two subsections. \ref{sec:results - biological view}: The analysis of some subtrees using 
      biological background knowledge. \ref{sec:results - cross validation}: Presentation of the results 
      of the leave-100-out cross validation.

@@

  We evaluated our ancestral state reconstruction and the relational state prediction using three 
    methods: a simulation, a selective analysis from a biological perspective and a leave-100-out 
    cross validation.

>>>>

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %--------------------------------------------------------------------------------------------------- 
  \section{Motivation}
    The aim of this thesis is the application of maximum parsimony algorithms to non-binary 
      trees and very large data sets. In particular, the example find the 'Origins of Parasitism' 
      throughout the Eukaryota Tree of Life. \\
    For these large phylogenetic synthesis trees, however, ancestral state reconstruction has so far 
      only been done for Bacteria and Archaea for binary traits by Goberna and Verdú \cite{Goberna2015}.
      However, this differs from eukaryotes in the sense that complex traits such as parasitism depend
      on more than one gene. \\
    \todo{Warum ist das relevant? Was könnte man dann tun?}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %--------------------------------------------------------------------------------------------------- 
  \section{Background}
    For about 50 years, people have been working on ancestral state reconstruction. The first paper 
      is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
      \cite{Camin1965}. Different methods have been developed and the question is which method is the 
      most suitable for the problem at hand: The ancestral state reconstruction for a huge 
      multifurcated tree with \colorbox{red}{binary/two discrete} states. \\
    Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
    The first is maximum parsimony: explain the current state with the least number of state changes 
      between the child and its ancestor. \\
    The other class she presents describes modeling the character evolution as a stochastic process and 
      using the likelihoods to compute the possible ancestral character states. This is generally done 
      with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

@@

    We will test the existing maximum parsimony algorithms of Fitch \cite{Fitch1971} and Sankoff 
      \cite{Sankoff1975} for this task and estimate their predictive power. The present tree structure 
      of OTL is not binary. A tree is multifurcated if each node has multiple ($n >= 3$) children 
      \cite{Felsenstein2003}. \\
    Maximum parsimony methods are developed for phylogenies, which are usually depicted as binary trees.
      Parsimonious in phylogeny refers to favoring the tree that needs the least evolutionary change 
      to explain the observed data. In our case, it is about the change of states 'is free-living' or 
      'is parasitic'. \\
      % Parsimonious in this context means to minimize the number of transitions between species states 
      % (free-living or parasitic). 
    The Sankoff method is implemented by Louca et al. for the non-binary case and is available as an R 
      package called \textit{Castor} \cite{Louca2017}. In addition, we have implemented the Fitch 
      method and adapted it for multifurcated trees. \\% to make it usable and comparable for our current problem.
    This achieves that we can predict ancestral states and unknown states of living species for large 
      non-binary relatives trees. \\

@@

%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%---------------------------------------------------------------------------------------------------
%--------------------------------------------------------------------------------------------------- 
\chapter{Methods}
  As stated in the introduction, in this thesis, a maximum parsimony algorithm is applied to the 
    whole tree of life to obtain an ancestral state reconstruction of free-living versus parasite 
    states. \\
  So far, these reconstructions have been made mainly on binary trees with better data availability. 
    Therefore, a simulation is first performed to evaluate existing algorithms and decide how they 
    may be adapted to our given problem. This is to perform the ancestral state reconstruction for a 
    multifurcated (non-binary) tree using binary states. \\
  Accordingly, in addition to the necessary data sets (GloBI, OTL), the chosen algorithm and the 
    evaluation of its results, this chapter also deals with the previously performed simulation and 
    the evaluation of the various algorithms and their parameters. \\
  Figure \ref{fig:workflow} briefly outlines these relationships. A more detailed view of the 
    workflow can be found in the appendix \ref{fig:BigWorkflow}.
  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\textwidth]{Figures/Workflow-overview.png}
    \caption{The Workflow of the resulting procedure with the following steps: \\
      (1) Retrieve phylogenentic tree data as input for the tree (OTL) and the state information (GloBI).
      (2) Get metadata of these for a realisitc simulation of the maximum parsimony algorithms (Fitch \& Sankoff).
      (3) Build and run the simulation.
      (4) Evaluation of parameters for the simulation and the ancestral state reconstruction of the real tree.
      (5) Evaluate the accuracy of developed algorithms and choose the best.
      (6) Run the resulting algorithm on the original data.
      (7) Evalute and interprete the results. %$\rightarrow$ Origins etc...
    }
    \label{fig:workflow}
  \end{figure}

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Description of data sets}
    Two types of data are needed for an ancestral state reconstruction: a tree and information about 
      the states. \\
    For the tree Open Tree of Life (OTL) is used \cite{Hinchliff2015} and for the state information 
      the Global Biotic Interaction database (GloBI) is taken \cite{Poelen2014}. \\
    % These are presented below.

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 0mm, clip, width=0.45\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{Simulated distributions of the states: \\
          60~\% Free-living - 40~\% Parasites \\
          red: parasites, blue: free-living, \\
          the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

@@

    Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
      use case here. Following the principle of the simpler model first. \\
    \anmerkungstext{Nochmal darstellen warum und selbst wenn sie fully sufficient wäre, heißt es ja 
      nicht, dass man sie nehmen muss. Also: motiviere! (Renard)} \\
    Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
      previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
      \cite{Farris1970}): Fitch parsimony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
    \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
      Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
    \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

      Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
        a set of states. There are only two states in this thesis, free-living (FL) and parasitic (P). 
        Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
        has only one. Leaf nodes have their state as a set (e.g. \{FL\} or \{P\}, unknown leaf nodes the 
        union of all possible states (\{FL, P\}). \\

      The algorithm traverses three times through the tree and fills these sets. \\
      In each step, two sets are considered and their intersect formed. There are two cases:
      \begin{enumerate}
        \item The intersection is not empty and corresponds to the new set.
        \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
      \end{enumerate}
      \begin{wrapfigure}{r}{0.5\textwidth}
        \centering
        \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
        \caption{Fitch algorithm for binary trees. \\
          The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
          root node) consists of three sets, where the last set is the final one (bold). \\
          From the second internal node (seen from the root node) there are several possibilities to 
          create the second and third set.}
        \label{fig: binary Fitch}
      \end{wrapfigure}
      First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Real data analysis}
    \todo{Evaluation section in methods?}
    \begin{itemize}
      \item Import tree
      \item Import interactions
      \item run castor algorithm / and others?
      \item interprete results (leave-100-out cross-validation)
    \end{itemize}
    We then evaluated the results in two ways. On the one hand, we have examined the results of some 
      subtrees for their credibility with the help of biological background knowledge. On the other 
      hand, we performed a leave-100-out cross-validation.

@@

  We then evaluated the results in two ways. On the one hand, we have examined the results of some 
    subtrees for their credibility with the help of biological background knowledge (subsection 
    \ref{sec:results - biological view}). On the other hand, we performed a leave-100-out cross-validation
    (Subsection \ref{sec:results - cross-validation}) and predicted approximately 98.17~\% states 
    correct.

@@

    We also have for a first overview we collected for each node its number of children (degree $-1$), 
      and plotted this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
    \begin{figure}
      \centering
      \begin{subfigure}[b]{0.59\textwidth}
        \includegraphics[width=0.9\textwidth]{Figures/multifurc.pdf}
        \caption{Histogram with automatic binwidth. \\ ~}
      \end{subfigure}
      \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[trim = 0mm 0mm 30mm 0mm, clip, width=0.9\textwidth]{Figures/multifurc_small.pdf}
        \caption{Histogram with $binwidth = 1$. Higher multifurcation than 30 has been cut off. \\ light blue: binary; \\ dark blue: multifurcation}
      \end{subfigure}
      \caption{Histograms about the multifurcation of the internal nodes of the synthesis tree. \\
        For each node, the number of children (degree $-1$) was collected. A node is multifurcated 
        if it has more than two children, so we deducted two from each number of children. We have 
        plotted this in two histograms: number of children $-2$ on the x-axis with log scale and the 
        number of nodes with this amount on the y-axis.}
      \label{fig:childrenOfNodes}
    \end{figure}
    The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
      every node (a node with two children is binary). That means it describes the number of nodes 
      which we have lost from the real (binary) phylogenetic tree. \\
    It can be recognized that we are very far from a binary tree. \\

@@

      \begin{table}[h]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            \bfseries Model / Taxa & \bfseries Kingdom & \bfseries Phylum & \bfseries Class & \bfseries Order \\
            \hline \hline
            multifurc $\sim$ taxa & 8273333 & \cellcolor{green!15}7937828 & \cellcolor{green!20}7842157 & \cellcolor{green!30}7644249 \\
            multifurc $\sim$ taxa & 8257680 & \cellcolor{green!15}7922207 & \cellcolor{green!20}7826490 & \cellcolor{green!35}7574154 \\
            \hline
            multifurc $\sim$ taxa + depth & 8273318 & \cellcolor{green!15}7934322 & \cellcolor{green!20}7839364 & \cellcolor{green!35}7539999 \\
            multifurc $\sim$ taxa + max.height & \cellcolor{green!15}7993515 & \cellcolor{green!25}7749121 & \cellcolor{green!30}7661817 & \cellcolor{green!40}7416211 \\
            multifurc $\sim$ taxa + min.height & 8251211 & \cellcolor{green!20}7875521  & \cellcolor{green!25}7778327 & \cellcolor{green!35}7516883 \\
            multifurc $\sim$ taxa + mean.height & \cellcolor{green!20}7825417 & \cellcolor{green!30}7644249 & \cellcolor{green!35}7572474 & \cellcolor{green!45}7340741 \\
            \hline
            multifurc $\sim$ taxa * depth & 8235932 & \cellcolor{green!20}7836755 & \cellcolor{green!25}7757688 & \cellcolor{green!45}7383808 \\
            multifurc $\sim$ taxa * max.height & \cellcolor{green!15}7963438 & \cellcolor{green!30}7693555 & \cellcolor{green!30}7614820 & \cellcolor{green!45}7335338 \\
            multifurc $\sim$ taxa * min.height & 8214030 & \cellcolor{green!20}7808940 & \cellcolor{green!30}7690618 & \cellcolor{green!45}7336627\\
            multifurc $\sim$ taxa * mean.height & \cellcolor{green!25}7768360 & \cellcolor{green!35}7536296 & \cellcolor{green!50}7484953 & \cellcolor{green!50}7206369 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BIC (Bayesian information criterion) values of the multifurcation models. \\
          These models were created with the R function \textit{glm()} and compared with the 
            \textit{BIC()} function. This results in the listed BIC values.}
        \label{table:BIC multifurcation} 
      \end{table}
      % * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\
      Within the every complexity class it can be seen that the mean height gives the best additional factor. \\
      Despite higher complexity, the BIC values are getting smaller from model to model, meaning that 
        the finest model available here is also the best one of these. Lower taxa than orders (e.g. 
        family) were computationally too expensive to calculate. \\
      The model \textit{multifurc $\sim$ order * mean.height} turns out to be the best of our models, 
        whereby it is possible that e.g. \textit{multifurc $\sim$ family * mean.height} is better.

@@

      Next to the problem of the multifurcation of the tree is the less of interaction data we have for 
        the species. For the ancestral state reconstruction, we need information about the states 
        (free-living or parasite) in the leaf nodes. \\
      The eukaryotic synthesis tree has 293,463 leaf nodes. The GloBI database has 5,346,414 interactions 
        (at 29.01.2018). Out of this data we got 51,337 distinct free-living species and 47,332 
        distinct parasite species for the whole tree of life. From the Eukaryota we could determine 
        25,992 and 34,879 species as parasites and free-living. With 2,293,463 leaf nodes we still have 
        about 97.34~\% unknown leaf nodes. In the discussion, section 
        \ref{sec:discussion - data situation}, we will talk about ways to extend this data. \\

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            \bfseries Model / Taxa & \bfseries Kingdom & \bfseries Phylum & \bfseries Class & \bfseries Order \\
            \hline \hline
            multifurc $\sim$ taxa & \cellcolor{green!15}545799 & \cellcolor{green!35}500004 & \cellcolor{green!45}485121 & \cellcolor{green!45}484681 \\
            \hline
            multifurc $\sim$ taxa + depth & \cellcolor{green!15}544862 & \cellcolor{green!40}493808 & \cellcolor{green!45}481869 & \cellcolor{green!50}478851 \\
            \hline
            multifurc $\sim$ taxa * depth & \cellcolor{green!15}544179 & \cellcolor{green!45}489845 & \cellcolor{green!45}481494 & \cellcolor{green!50}478188 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BIC (Bayesian information criterion) values unknown state information models. \\
          These models were created with the R function \textit{glm()} and compared with the 
            \textit{BIC()} function. This results in the listed BIC values.}
        \label{table:BIC unknown information} 
      \end{table}
      It also follows from this table that the most complex model is the best. In general, the BIC 
        values are smaller than those of the multifurcation models. The modeling here is thus better. \\
      Again, the calculation of finer models (e.g. order or family) was too expensive. \\

@@

    As presented, we compare two methods in our simulation to their prediction accuracy: Fitch and 
      Sankoff. \\
    Figure \ref{fig:influence of unknown data} shows the results of examining various parameters.
    \begin{figure}[h!]
      \centering
      \includegraphics[trim = 0mm 100mm 0mm 0mm, clip, width=\textwidth]{Figures/simulation_evaluation_1.png}
    \end{figure}
    ~ \\ ~ \\ ~ \\
    \begin{figure}
      \centering
      \includegraphics[trim = 0mm 0mm 0mm 150mm, clip, width=\textwidth]{Figures/simulation_evaluation_1.png}
      \caption{Influence of unknown data to prediction. \\
        The first column describes the distributions of free-livings and parasites with a given 
          threshold for the respective simulations to the right. \\
        The middle column investigates the influence of the unknown states, the right the influence of
          the strength of the multifurcation. \\
        The y-axes indicate the percentage of correctly predicted states (including known states). On 
          the x-axis the percentage of forgotten states or missing internal nodes. \\
        Each point corresponds to the average of one hundred simulations, each with 10,000 leaf nodes. \\
        For the middle column we set the strength of the multifurcations to 0.95~\% similar to the 
          real data and in the right column the amount of the unknowns to 0.95~\% also similar to the 
          real data.}
      \label{fig:influence of unknown data}
    \end{figure}
    First, we tested different transition probability distributions including threshold (first column). 
      It can be observed that the more balanced the percentages of free-livings and parasites, the 
      worse the prediction of the algorithms (second and third column). \\
    It can be seen that both algortithms are always over 50~\% prediction and therefore better than 
      guessing. Moreover, they are usually close to each other, with Sankoff always makes better 
      predictions except for equally distributed states as Fitch. \\
    On the other hand, we examined the influence of missing internal nodes (ridge of multifurcation) 
      and missing leaf node information (unknown leaf nodes). \\
    Both factors have a relatively linear influence on the Sankoff method. Fitch, on the other hand, 
      breaks significantly in his prediction from about 70\% unknown leaf nodes or mutifurcation. \\

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Results of the real data analysis created with Sankoff}
    This section is about evaluating the prediction of real data using the Sankoff method. \\
    It is divided into two subsections. \ref{sec:results - biological view}: The analysis of some subtrees using 
      biological background knowledge. \ref{sec:results - cross-validation}: Presentation of the results 
      of the leave-100-out cross-validation.

@@

  We evaluated our ancestral state reconstruction and the relational state prediction using three 
    methods: a simulation, a selective analysis from a biological perspective and a leave-100-out 
    cross-validation.
